An Empirical Analysis of the Role of Energy in Economic Growth

### 2013-03-28 EconGrowth-R-Functions.Rnw
Added a 2-questions slide after the CD model in the presentation. Also trying (unsuccessfully) to add a point to the 3-D SF graph.

#########################
# Creates a 3-D wireframe graph with SSE on the vertical axis and 
# lambda and m on the horizontal axes.
# params: 
#   countryAbbrev: 2-letter country code (upper case)
#   factor: the factor of production you want to use in the single-factor model.
##
sf3DSSEGraph <- function(countryAbbrev, factor){
  # Generate the data in the form of a matrix with gamma in columns, lambda in rows, and SSE in the middle
  # First, make a data.frame with which we want to make predictions
  lambdaSeq <- seq(0.0, 0.05, by=0.01)
  mSeq <- seq(0.0, 1.0, by=0.1)

@@

@@

  # Create the lattice graph
  fig <- wireframe(sse ~ lambda * m,
                   data = dataFor3DGraph,
                   screen = list(z = -20, x = -55),
                   xlab = "$\\lambda$",
                   ylab = "$\\gamma$",
                   zlab = list(label="$SSE_y$", rot=90),
                   scales = list(arrows=FALSE, 
                                 cex=0.45, 
                                 col="black", 
                                 tck=0.8,
                                 x=list(at=c(0.00, 0.01, 0.02, 0.03)),
                                 y=list(at=c(0.0, 0.5, 1.0)),
                                 z=list(at=c(10, 20, 30))
                                 ),
                   xlim = c(0.0, 0.04), # lambda axis
                   ylim = c(0.0, 1.0), # gamma axis
                   zlim = c(0, 30) # sse axis
                   )
  return(fig)    
}
@

>>>>

#########################
# Creates a 3-D wireframe graph with SSE on the vertical axis and 
# lambda and m on the horizontal axes.
# params: 
#   countryAbbrev: 2-letter country code (upper case)
#   factor: the factor of production you want to use in the single-factor model. One of K, L, Q, X, or U
##
sf3DSSEGraph <- function(countryAbbrev, factor){
  # Generate the data in the form of a matrix with gamma in columns, lambda in rows, and SSE in the middle
  # First, make a data.frame with which we want to make predictions
  lambdaSeq <- seq(0.0, 0.05, by=0.01)
  mSeq <- seq(0.0, 1.0, by=0.1)

@@

  # Get information about the optimum point, the point where SSE is minimized
  coefs <- coef(model)
  lambda_opt <- coefs["lambda"]
  gamma_opt <- coefs["gamma"]
  #sse_opt <- ???????

@@

  # Create the lattice graph
#   fig <- wireframe(sse ~ lambda * m,
#                    data = dataFor3DGraph,
#                    screen = list(z = -20, x = -55),
#                    xlab = "$\\lambda$",
#                    ylab = "$\\gamma$",
#                    zlab = list(label="$SSE_y$", rot=90),
#                    scales = list(arrows=FALSE, 
#                                  cex=0.45, 
#                                  col="black", 
#                                  tck=0.8,
#                                  x=list(at=c(0.00, 0.01, 0.02, 0.03)),
#                                  y=list(at=c(0.0, 0.5, 1.0)),
#                                  z=list(at=c(10, 20, 30))
#                                  ),
#                    xlim = c(0.0, 0.04), # lambda axis
#                    ylim = c(0.0, 1.0), # gamma axis
#                    zlim = c(0, 30) # sse axis
#                    )

  fig <- wireframe(sse ~ lambda * m,
                   data = dataFor3DGraph,
                   screen = list(z = -20, x = -55),
                   xlab = "$\\lambda$",
                   ylab = "$\\gamma$",
                   zlab = list(label="$SSE_y$", rot=90),
                   scales = list(arrows=FALSE, 
                                 cex=0.45, 
                                 col="black", 
                                 tck=0.8,
                                 x=list(at=c(0.00, 0.01, 0.02, 0.03)),
                                 y=list(at=c(0.0, 0.5, 1.0)),
                                 z=list(at=c(10, 20, 30))
                                 ),
                   xlim = c(0.0, 0.04), # lambda axis
                   ylim = c(0.0, 1.0), # gamma axis
                   zlim = c(0, 30) # sse axis
                   )
  return(fig)    
}
@

### 2013-03-28 EconGrowthSeminar.Rnw
Added a 2-questions slide after the CD model in the presentation. Also trying (unsuccessfully) to add a point to the 3-D SF graph.

% Some commands for parameter tables in slides
\newcommand{\yRow}{$y$ & $Y/Y_0$, output}
\newcommand{\timeRow}{$t$ & year}
\newcommand{\initialTimeRow}{$t_0$ & initial year}
\newcommand{\kRow}{$k$ & $K/K_0$, capital stock}
\newcommand{\lRow}{$l$ & $L/L_0$, labor}
\newcommand{\eRow}{$e$ & $E/E_0$, one of $q$, $x$, or $u$}
\newcommand{\lambdaRow}{$\lambda$ & technological progress}
\newcommand{\alphaRow}{$\alpha$ & ``strength'' of capital}
\newcommand{\betaRow}{$\beta$ & ``strength'' of labor}
\newcommand{\gammaRow}{$\gamma$ & ``strength'' of energy}

@@

\begin{frame}{Type of Energy ($E$)} %----------------------------
  \begin{itemize}
    \item{Thermal energy ($Q$)}
    \vspace{\baselineskip}  
    \item{Exergy ($X$)}
    \vspace{\baselineskip}
    \item{Useful work ($U$)}
  \end{itemize}
  \vspace{\baselineskip}
  \pause
  \begin{center}Simplification: 1 type of energy ($Q$) for this presentation\end{center}
\end{frame}

@@

    \begin{tabular}{|c|l|}\hline
      \yRow\\
      \timeRow\\
      \initialTimeRow\\
      $f$ & $F$/$F_0$, a factor of production\\
      \hline
      \lambdaRow\\
      $m$ & ``strength'' of factor $f$\\
      \hline
    \end{tabular}
  \end{center}
\end{frame}

@@

    \begin{itemize}
      \item{Time-dependent technology term: $\mathrm{e}^{\lambda(t-t_0)}$}
      \begin{itemize}
        \item{Represents the irrepressible ``progress'' of technology}
        \item{Is independent of factor of production ($f$)}
      \end{itemize}
      \item{Use for illustration purposes here}
      \item{Example: \citet{Nel:2009wq} use $f$ = $q$}
      \item{Parameters estimated by \texttt{nls} function in \texttt{R}}
    \end{itemize}
\end{frame}

@@

\begin{frame}{$SSE$ Minimization}{US single-factor with $Q$} %---------------------------- 
<<3D_Graph_US_SF_Q, fig.width=presentationGraph1ColWidth*1.3, fig.height=presentationGraph1RowHeight*1.3>>=
sf3DSSEGraph("US", "Q")
@
  % Including at bottom of frame so as not to push other material downward in the frame.
  \putFlagUpperRight{US}{0.055}  
\end{frame}

<<Coefs_US_SF>>=
sfUSK <- coef(singleFactorModel("US", "K"))
sfUSL <- coef(singleFactorModel("US", "L"))
sfUSQ <- coef(singleFactorModel("US", "Q"))
sfUSX <- coef(singleFactorModel("US", "X"))
sfUSU <- coef(singleFactorModel("US", "U"))
@
\begin{frame}{Single-Factor $y$ Predictions} %----------------------------
  \begin{columns}[c]
    \column{2.2in}
      \begin{center}
        \sfEqn
        \vspace{\baselineskip}

@@

\begin{frame}{Comments on Cobb-Douglas Model Equation} %----------------------------
  \begin{center} \cdeEqn \end{center}
  \begin{itemize}
    \item{Time-dependent technology term: $\mathrm{e}^{\lambda(t-t_0)}$}
    \begin{itemize}
      \item{Represents the irrepressible ``progress'' of technology}
      \item{Is independent of factors of production ($k$, $l$, and $e$)}
    \end{itemize}
    \item{$k$, $l$, and $e$ factors of production ``blend'' with $\alpha + \beta + \gamma = 1$}
    \item{Example: \citet{Froling:2009vo}}
  \end{itemize}
\end{frame}

@@

@@

    \begin{tabular}{|c|l|}\hline
      \yRow\\
      \timeRow\\
      \initialTimeRow\\
      \kRow\\
      \lRow\\
      \eRow\\
      \hline
      $\gamma$   &   fitting coefficient\\
      \lambdaRow\\
      $\delta_1$   &   split between ($k$) and ($l$)\\
      $\rho_1$     &   ``strength'' of ($k$) vs. ($l$)\\
      $\delta$   &   split between ($kl$) and ($e$)\\
      $\rho$     &   ``strength'' of ($kl$) vs. ($e$)\\
      \hline
    \end{tabular}
  \end{center}
\end{frame}

\begin{frame}{Comments on CES Model Equation} %----------------------------
  \begin{center} \small \cesEqn \normalsize \end{center}
  \begin{itemize}
    \item{Time-dependent technology term: $\mathrm{e}^{\lambda(t-t_0)}$}
    \begin{itemize}
      \item{Represents the irrepressible ``progress'' of technology}
      \item{Is independent of factors of production ($k$, $l$, and $e$)}
    \end{itemize}
    \item{$\delta$: split between ($kl$) and ($e$)}
    \item{$\sigma$: elasticity of substitution ($0.0 \leq \sigma \leq 1.0$)}
    \begin{itemize}
      \item{$\sigma$ ******* rightarrow******* 0: ($kl$) and ($e$) are complementary}
      \item{$\sigma$ ******* rightarrow******* 1: ($kl$) and ($e$) are substitutes}
    \end{itemize}
    \item{Example: \citet{Stern:2012ey}}
    \item{Parameters estimated by \texttt{cesEst} function in \texttt{R}'s \texttt{micEcon} package}
  \end{itemize}
\end{frame}

>>>>

% Some commands for parameter tables in slides
\newcommand{\yRow}{$y$ & $Y/Y_0$, output}
\newcommand{\timeRow}{$t$ & year}
\newcommand{\initialTimeRow}{$t_0$ & initial year}
\newcommand{\kRow}{$k$ & $K/K_0$, capital stock}
\newcommand{\lRow}{$l$ & $L/L_0$, labor}
\newcommand{\eRow}{$e$ & $E/E_0$, one of $q$, $x$, or $u$}
\newcommand{\lambdaRow}{$\lambda$ & technological progress}
\newcommand{\alphaRow}{$\alpha$ & ``influence'' of capital}
\newcommand{\betaRow}{$\beta$ & ``influence'' of labor}
\newcommand{\gammaRow}{$\gamma$ & ``influence'' of energy}

@@

\begin{frame}{Type of Energy ($E$)} %----------------------------
  \begin{itemize}
    \item{Thermal energy ($Q$)}
    \vspace{\baselineskip}  
    \item{Exergy ($X$)}
    \vspace{\baselineskip}
    \item{Useful work ($U$)}
  \end{itemize}
  \vspace{\baselineskip}
  \pause
  \begin{center}Simplification: One type of energy ($Q$) for this presentation\end{center}
\end{frame}

@@

    \begin{tabular}{|c|l|}\hline
      \yRow\\
      \timeRow\\
      \initialTimeRow\\
      $f$ & $F$/$F_0$, a factor of production\\
      \hline
      \lambdaRow\\
      $m$ & ``influence'' of factor $f$\\
      \hline
    \end{tabular}
  \end{center}
\end{frame}

@@

    \begin{itemize}
      \item{$\mathrm{e}^{\lambda(t-t_0)}$: Time-dependent technology term}
      \begin{itemize}
        \item{Represents the irrepressible ``progress'' of technology}
        \item{Is independent of factor of production ($f$)}
      \end{itemize}
      \item{Used for illustration purposes here}
      \item{Example: \citet{Nel:2009wq} use $f$ = $q$}
      \item{Parameters estimated by \texttt{nls} function in \texttt{R}}
    \end{itemize}
\end{frame}

@@

\begin{frame}{$SSE$ Minimization}{US single-factor with $Q$} %---------------------------- 
<<3D_Graph_US_SF_Q, fig.width=presentationGraph1ColWidth*1.3, fig.height=presentationGraph1RowHeight*1.3, eval=TRUE>>=
sf3DSSEGraph("US", "Q")
@
  % Including at bottom of frame so as not to push other material downward in the frame.
  \putFlagUpperRight{US}{0.055}  
\end{frame}

<<Coefs_US_SF>>=
sfUSK <- coef(singleFactorModel("US", "K"))
sfUSL <- coef(singleFactorModel("US", "L"))
sfUSQ <- coef(singleFactorModel("US", "Q"))
sfUSX <- coef(singleFactorModel("US", "X"))
sfUSU <- coef(singleFactorModel("US", "U"))
@
\begin{frame}{Single-factor (with $q$) Predictions for $y$} %----------------------------
  \begin{columns}[c]
    \column{2.2in}
      \begin{center}
        \sfEqn
        \vspace{\baselineskip}

@@

\begin{frame}{Comments on Cobb-Douglas Model Equation} %----------------------------
  \begin{center} \cdeEqn \end{center}
  \begin{itemize}
    \item{$\mathrm{e}^{\lambda(t-t_0)}$: Time-dependent technology term}
    \begin{itemize}
      \item{Represents the ``irrepressible'' ``progress'' of technology}
      \item{Is independent of factors of production ($k$, $l$, and $e$)}
    \end{itemize}
    \item{$k$, $l$, and $e$ factors of production ``blend'' with $\alpha + \beta + \gamma = 1$}
    \item{Example: \citet{Froling:2009vo}}
  \end{itemize}
\end{frame}

@@

\begin{frame}{Two Important Questions for\\Cobb-Douglas Model} %----------------------------
  \begin{center}
    \cdeEqn
  \end{center}
  \begin{itemize}
    \item{Is energy important for this economy?\\Yes, if $\gamma$ not small.}
    \vspace{\baselineskip}

    \item{Are we sure? (Is energy statistically significant?)\\Yes, if $CI_{\gamma}$ doesn't cross 0.0.}
  \end{itemize}
\end{frame}

@@

    \begin{tabular}{|c|l|}\hline
      \yRow\\
      \timeRow\\
      \initialTimeRow\\
      \kRow\\
      \lRow\\
      \eRow\\
      \hline
      $\gamma$   &   fitting coefficient\\
      \lambdaRow\\
      $\delta_1$   &   split between ($k$) and ($l$)\\
      $\rho_1$     &   ``influence'' of ($k$) vs. ($l$)\\
      $\delta$   &   split between ($kl$) and ($e$)\\
      $\rho$     &   ``influence'' of ($kl$) vs. ($e$)\\
      \hline
    \end{tabular}
  \end{center}
\end{frame}

\begin{frame}{Comments on CES Model Equation} %----------------------------
  \begin{center} \small \cesEqn \normalsize \end{center}
  \begin{itemize}
    \item{$\mathrm{e}^{\lambda(t-t_0)}$: Time-dependent technology term}
    \item{$\delta$: split between ($kl$) and ($e$)}
    \item{$\sigma \equiv \frac{1}{1+\rho}$: elasticity of substitution ($0 \leq \sigma \leq 1$)}
    \begin{itemize}
      \item{$\sigma \rightarrow 0$: ($kl$) and ($e$) are complementary}
      \item{$\sigma \rightarrow 1$: ($kl$) and ($e$) are substitutes}
    \end{itemize}
    \item{Example: \citet{Stern:2012ey}}
    \item{Parameters estimated by \texttt{cesEst} function in \texttt{R}'s \texttt{micEcon} package}
  \end{itemize}
\end{frame}
